---
title: "Unsupersived Learning - K Means Clustering and PCA"
author: "Eliud Mule"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set working Directory
```{r}
# setwd("C:/Users/user/Desktop/Professional DS Mule/R")
```


## **Task 1 : Data Import and Exploration**
### **Load required libraries**
```{r}
library(tidyverse)      # Data Manipulation
library(cluster)          # Clustering Algorithms
library(factoextra)       # Clustering Visualization
library(palmerpenguins)      # Sample Dataset
library(dplyr)            # Data Manipulation
library(ggplot2)          # Data Visualization

```


### **Load Penguin Dataset**
```{r}
data("penguins", package = "palmerpenguins")
head(penguins)
```

### **Explore the dataset**
```{r}
library(psych)
str(penguins)
describe(penguins)
```

### **Check for missing values**
```{r}
# Count missing values per column
missing_count <- colSums(is.na(penguins))
missing_count

```

```{r}
# percentage of missing values
sum(is.na(penguins)) / (nrow(penguins) * ncol(penguins)) * 100
```

### **remove rows with missing values**
```{r}
penguins_clean <- na.omit(penguins)
dim(penguins_clean)
```

## **Task 2 :Data Preprocessing**
```{r}
# Select only numeric columns for clustering
numeric_vars <- c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g")
penguins_numeric <- dplyr::select(penguins_clean,all_of(numeric_vars))

```


```{r}
# Scale variables for clustering
penguins_scaled <- as.data.frame(scale(penguins_numeric))
```

```{r}
# Verify Scaling
colMeans(penguins_scaled)  # Should be close to 0
apply(penguins_scaled, 2, sd)  # Should be close to 1
```


## **Task 3 : Exploratory Data Analysis**
```{r}
# Pairs plot colored by species
pairs(penguins_numeric, col = penguins_clean$species, pch = 19,cex = 0.7, main = "Penguin Morphology : Pairs Plot")


```




```{r}
# Boxplots for each numeric variable by species
penguins_clean %>%
  pivot_longer(cols = all_of(numeric_vars), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = species, y = value, fill = species)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Boxplots of Numeric Variables by Species") +
  theme_minimal() +
  theme(legend.position = "none")
```



```{r}
# Correlation matrix and heatmap
cor_matrix <- cor(penguins_numeric)
cor_matrix

```



```{r}

corrplot::corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black", tl.srt = 45, title = "Correlation Matrix Heatmap")

```


## **Task 4 : K-Means Clustering**
```{r}
set.seed(2026)  # For reproducibility
# Determine optimal number of clusters using Elbow Method
fviz_nbclust(penguins_scaled, kmeans, method = "wss") +
  labs(title = "Elbow Method for Optimal Clusters")

```

```{r}
# Determine optimal number of clusters using Silhouette Method
fviz_nbclust(penguins_scaled, kmeans, method = "silhouette") +
  labs(title = "Silhouette Method for Optimal Clusters")
```


From the Elbow and Silhouette methods, we can see that the optimal number of clusters is 2.


## **Task 5 : K-Means : Model Fitting and Cluster Profiling**
```{r}
set.seed(2026)  # For reproducibility
k_opt <- 2
# Fit K-Means model with 3 clusters
kmeans_model <- kmeans(penguins_scaled, centers = k_opt, nstart = 25)
kmeans_model

```

```{r}
# Add cluster assignments to original data
penguins_clean$Cluster <- as.factor(kmeans_model$cluster)
#Cluster centroids
kmeans_model$centers
# Cluster profiling
cluster_profiles <- penguins_clean %>%
  group_by(Cluster) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE),n = n())
cluster_profiles

```




```{r}
# Table summarizing cluster characteristics(size,centroid values,dominant species)
cluster_summary <- penguins_clean %>%
  group_by(Cluster) %>%
  summarise(Size = n(),
            Centroid_Bill_Length = mean(bill_length_mm),
            Centroid_Bill_Depth = mean(bill_depth_mm),
            Centroid_Flipper_Length = mean(flipper_length_mm),
            Centroid_Body_Mass = mean(body_mass_g),
            Dominant_Species = names(sort(table(species), decreasing = TRUE)[1]))
cluster_summary


```





```{r}
# Visualization
fviz_cluster(kmeans_model, data = penguins_scaled, geom = "point", ellipse.type = "norm", main = "K-Means Clustering of Penguins")
```


## **Task 6 : Principal Component Analysis (PCA)**
```{r}
# Perform PCA
pca_result <- prcomp(penguins_scaled, center = TRUE, scale. = TRUE)
# Summary of PCA
summary(pca_result)

```

```{r}
# Scree plot to visualize explained variance
fviz_eig(pca_result, addlabels = TRUE, main = "Scree Plot : PCA Explained Variance")
```


```{r}
# Loadings of the first two principal components
pca_result$rotation[, 1:2]
```




```{r}
# Biplot of the first two principal components
fviz_pca_biplot(pca_result, label = "var", habillage = "none", addEllipses = FALSE,col.var = "black", repel = TRUE,title = "PCA Biplot of Penguins : PC1 vs PC2")

```


## **Task 7 : Combining K -means and PCA**
```{r}
set.seed(2026)
# Project data onto the first two principal components
pca_scores <- as.data.frame(pca_result$x[, 1:2])
colnames(pca_scores) <- c("PC1", "PC2")

#K-means on PCA- reduced data
kmeans_pca <- kmeans(pca_scores, centers = k_opt, nstart = 25)
# Compare cluster assignments from PCA -reduced data  on original4-dimensional data using confusion matrix
table(FullSpace = kmeans_model$cluster, PCspace = kmeans_pca$cluster)

```

```{r}
# Scatterplot of PC1 vs PC2 colored by K-Means clusters(full data)
pca_scores$Cluster <- as.factor(kmeans_model$cluster)
ggplot(pca_scores, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "PC1 vs PC2 :Colored by K-Means Clusters (Full Data)") +
  theme_minimal() +
  scale_color_manual(values = c("red", "green", "blue"))
```

```{r}
# Scatterplot of PC1 vs PC2 colored by species
pca_scores$Species <- penguins_clean$species
ggplot(pca_scores, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "PC1 vs PC2 : Colored by Species") +
  theme_minimal() +
  scale_color_manual(values = c("orange", "purple", "cyan"))
```


## **Task 8 :Cluster validation and Quality Assessment**
```{r}
# Silhouette analysis for K-Means clusters
silhouette_kmeans <- silhouette(kmeans_model$cluster, dist(penguins_scaled))
mean_silhouette_kmeans <- mean(silhouette_kmeans[, 3])
mean_silhouette_kmeans
plot(silhouette_kmeans, main = "Silhouette Plot for K-Means Clusters", col = c("red", "green", "blue"), border = NA)
```

```{r}
# Dunn index for K-Means cluster
library(clValid)
dunn_index <- dunn(dist(penguins_scaled), kmeans_model$cluster)
dunn_index
```


```{r}
# Calinski-Harabasz Index for K-Means clusters
library(clusterSim)

ch_index <- index.G1(penguins_scaled, kmeans_model$cluster)

ch_index
```


```{r}
# Comparing CH values for k= 2, 3, 4
ch_values <- numeric()

for (k in 2:4) {
  km <- kmeans(penguins_scaled, centers = k, nstart = 25)
  ch_values[k] <- index.G1(penguins_scaled, km$cluster)
}

ch_values
```

## **Task 9 : Interpretation and Biological Insights**
```{r}
#Contigency table of K-Means clusters vs species
table(penguins_clean$Cluster, penguins_clean$species)
```


```{r}
# PCA loadings Interpretation
loadings_df <- as.data.frame(pca_result$rotation[, 1:2])
loadings_df$Variable <- rownames(loadings_df)
print(loadings_df)
```




















